Fast Sorting on Modern Computers

### 2005-08-30 docs/tech_report/conclusions.tex
tidied up

We found that insertion sort has the best performance of any of the elementary
sorts, with fewer instructions, cache misses and exactly one branch
misprediction per key. Selection sort also has very few branch mispredictions,
and our results match Knuth's calculations. We find that reducing the number of
runs in bubblesort and shakersort at the cost of additional instructions is
counter productive, and that their branch prediction performance is incredibly
poor.

@@

The cache optimized heaps perform better than the non-optimized versions. At the
cost of a slightly increased instruction count, the number of branch misses can
be greatly reduced, and the overall performance of our 8-heap is significantly
better than the unoptimized version. This is despite our finding of that
increasing the fanout increases branch mispredictions.

Additionally, we created an improvement to heapsort which removed the need for a
sentinel, without introducing any more instructions.

@@

We found that algorithm S has a lower instruction count than algorithm N, but
that this does not necessarily translate to better performance. Our results show
also that the level 1 cache needs to be taken into account, that the extra
instruction in multi-mergesort make it slower than double-aligned tiled
mergesort, and that multi-mergesort has regular access patterns during the k-way
merge, which can be taken advantage of by a two-level adaptive predictor.

@@

\subsection{Elementary Sorts}
Insertion sort has exactly one miss per key. Using this result we were able to
make several other sorts faster by replacing their simple inline sorts with
insertion sorts.

Selection sort also has a very low number of branch mispredictions, the results
of which match up well to Knuth's calculation. We found as well that bubblesort
and shakersort have incredibly bad branch prediction performance. Both are
$O(N^2)$ sorts, and the number of branches and branch misses have the same
complexity.

TODO bubblesort

\subsection{Heapsort}
In heapsort, we found a pattern of branch mispredictions when using a fanout in
our 8-heap, and that this pattern matched closely that of selection sort,
discussed above. We found that the approximated results using $H_N$ was similar to
our expected results for large arrays, and followed a similar pattern for small
arrays.

@@

However, LaMarca's move to multi-quicksort, while not affecting the rate of
branch mispredictions in the partitioning phase, results in a large increase in
branch mispredictions due to the addition of a binary search. We determine that
this while reducing the instruction count, the increase is actual running time
due to branch mispredictions is significant.

@@

We are able to accurately predict the number of branch mispredictions in
shellsort, using our results from insertion sort. We also found that the
improvement mode to shellsort, separating the final loop from the rest, results
in keys being moved 1.07 places, up from 1.03, contributing to its improved
performance.

@@

Over the course of this research. The vast majority of our results have
correlated exactly with LaMarca's. Many of our graphs are exactly the same
shape to LaMarca's, can be calculated exactly or almost exactly using his
methods, or can be seen to be equivalent when scaled by an appropriate ratio
(due to different test parameters).

@@

However, we disagree with his results in several key places. Firstly, we note
that an in-lined sorting method does not perform as fast as insertion sort. We
find it difficult to perform his improvements on either algorithm N or algorithm
S, and write our own mergesort instead. We expected that mergesort would have
half the instruction count of heapsort we found instead they were very similar.

@@

\section{Cache Results}

% TODO more details in this
The test results in this report mostly agreed with LaMarca's results. The
heapsort graphs had nearly exactly the same shape as LaMarca's, and the ratios
between the base and cache-conscious versions were nearly identical. The
multi-mergesort results were slightly different: instruction count was
identical, but while the cache results of the base and tiled mergesorts were
identical, multi-mergesort results were not the same. However, the basis for
those results - a sharp increase in cache misses followed by a flat-lined graph
- were in both cases.

The quicksort results were also very similar, though the multi-quicksort results
had a very different shape to LaMarca's. However, the instruction count
results for each quicksort agreed with LaMarca's results, as did cache miss
results of the base and cache-conscious versions.

The two sets of radixsort results were slightly similar. The rise and plateau of
the curve to the right of the graphs was the same in both cases, though the left
hand side of the graphs did not agree.

Another interesting observation can be made: the results of the simulations on
the Pentium 4 were not always influenced by the cache misses. Though heapsort
and radixsort began to perform poorly once the data set no longer fit in the
cache, neither quicksort nor mergesort were affected. The reason why radixsort
was affected is obvious - there is no temporal reuse; in heapsort, the full
cache line is not being used, and spatial locality suffers as a result. It is
interesting that the other two sorts were unaffected, even though their cache
miss graphs indicate a significant increase in misses.

Finally, it was shown that in some cases a direct-mapped cache performed better
than a fully-associative one. The reasons for this were not clear.

\section{Branch Prediction Results}

Insertion sort has only one branch misprediction per key. This shows that it is
possible to predict comparative branches correctly, though the instruction cost
of such an algorithm may be high. This is also shown in multi-quicksort: though
the instruction count and memory reference count are both doubled, it is still
faster to use a sequential search than a binary search across small lists.

Selection sort has a similar number of instructions and cache misses as
bubblesort and a larger instruction and cache miss count than shakersort, yet it
still considerably outperforms both of these. This shows the high cost of branch
mispredictions on an algorithm.

Radixsort has almost no branch mispredictions at all. The only mispredictions
are due to its flow control, but it escapes comparative mispredictions, and this
factor allows it to perform as fast as, and in some cases faster than quicksort,
despite its high number of cache misses.

The heapsort results show the disadvantage of using unsorted lists. Even though
there are less branches in the 4-heap version, the number of branch
mispredictions are higher than in a standard 2-heap.

@@

\section{Best Sort}
The results also helped answer the question of which is the best sort. However,
it is important to remember that sorts have different purposes and properties,
so this shall be considered. A comparative chart of the fastest sorts is shown
in Figure \ref{All cycles}. This shows the cycles per key of each of the major
sorts and their derivatives, measured using Pentium 4 performance counters.
None of the elementary sorts or heapsorts are on the chart, as their results
are significantly worse than the other results, which makes it difficult to
accurately see them.

@@

This section lists the contributions of this report. Each of these is novel  
and was discovered or developed during the course of this report.
%TODO cant say that
The primary contribution of this report is that it repeats the work of LaMarca
on cache-conscious sorting, and validates many of his claims, while questioning
others. The next most important contribution is having performed a branch
prediction analysis of major sorts - which to my knowledge has not been done
before - and developing a framework by which these results can be repeated and
expanded upon.

>>>>

We found that insertion sort has the best performance of any of the elementary
sorts, with fewer instructions and cache misses than any other elementary sort,
and exactly one branch misprediction per key. Selection sort also has very few
branch mispredictions, and our results match figures created using Knuth's
harmonics formula. We found that bubblesort and shakersort have poor branch
prediction performance, and that this cripples their performance, especially
compared to selection sort which is otherwise similar in behaviour. We
determined that bubblesort has better branch performance at the start and end of
its sorting.

We also found that reducing the number of runs performed by shakersort and
bubblesort is counter productive due to additional book-keeping cost. We found
also that of the elementary sorts, only insertion sort has good cache
performance, as the others do not exploit any temporal locality.

@@

The cache optimized heaps perform better than the non-optimized versions. At the
cost of a slightly increased instruction count, the number of cache misses can
be greatly reduced, and the overall performance of our 8-heap is significantly
better than the unoptimized version. This is despite our finding of that
increasing the fanout increases branch mispredictions.

Additionally, we created an improvement to heapsort using a 2-heap, which
removed the need for a sentinel, without introducing any more instructions. We
found that a $k$-ary heap has similar branch prediction performance to selection
sort, though not exactly, in that the fanout becomes more predictable the larger
it is.

@@

We found that algorithm S has a lower instruction count than algorithm N, but
that this does not necessarily translate to better performance. Our results show
also that the level 1 cache needs to be taken into account, that the extra
instruction in multi-mergesort make it slower than tiled mergesort, and that
multi-mergesort has regular access patterns during the k-way merge, which can be
taken advantage of by a two-level adaptive predictor.

@@

Among them is the fact that bimodal branch predictors have performed better than
2-level adpative ones in almost every case in this research. This holds across
all comparative branches in all the sorts we surveyed, and we believe that where
explicit patterns - which can be exploited by 2-level adaptive predictors - do
not exist, a bimodal predictors perform better.

Insertion sort has exactly one miss per key. Using this result we were able to
make several other sorts faster by replacing their simple inline sorts with
insertion sorts.

Selection sort also has a very low number of branch mispredictions, the results
of which can be calculated using Knuth's formula for $H_N-1$. We found as well
that bubblesort and shakersort have incredibly bad branch prediction
performance. Both are $O(N^2)$ sorts, and the number of branches and branch
misses have the same complexity.

Bubblesort performs similarly to selection sort. It differs in that it partially
sorts the array as it moves the smallest key down the array, which selection sort
does not do. This has a very large negative effect for branch prediction
performance, which is not suffered by selection sort. This occurs in the middle
of the sort; at the start, it performs like selection sort, being randomly
arranged; at the end, it is largely sorted and is predictable again.

In heapsort, we found a pattern of branch mispredictions when using a fanout in
our 8-heap, and that this pattern matched closely that of selection sort,
discussed above. We found that though our approximated results using $H_N-1$
were not identical to our actual results - due to the arrays being very small -
our actual results did become more predictable in a similar fashion.

@@

However, LaMarca's move to multi-quicksort, while not affecting the rate of
branch mispredictions in the partitioning phase, results in a large increase in
branch mispredictions due to the addition of a binary search. We determine that
while this reduces the instruction count and number of memory accesses, the
increase is actual running time due to branch mispredictions is significant.

@@

We are able to accurately predict the number of branch mispredictions in
shellsort, using our results from insertion sort. We also found that the
improvement mode to shellsort, separating the final loop from the rest, results
in a significantly faster loop, with an extra speed boost from a side-effect
skipping an iteration with $h = 2$.

@@

Over the course of this research, the vast majority of our results have
correlated exactly with LaMarca's. Many of our graphs are exactly the same
shape to LaMarca's, can be calculated exactly or almost exactly using his
methods, or can be seen to be equivalent when scaled by an appropriate ratio.

@@

However, we disagree with his results in several key places. Firstly, we note
that an inlined sorting method does not perform as fast as insertion sort. We
find it difficult to perform his improvements on either algorithm N or algorithm
S, and write our own mergesort instead. We expected that mergesort would have
half the instruction count of heapsort we found instead they were very similar.

@@

\section{Branch Prediction Results}
%TODO mergre this section

@@

\section{Best Sort}

The results also helped answer the question of which is the best sort. However,
it is important to remember that sorts have different purposes and properties,
so this shall be considered. A comparative chart of the fastest sorts is shown
in Figure \ref{All cycles}. This shows the cycles per key of each of the major
sorts and their derivatives, measured using Pentium 4 performance counters.
None of the elementary sorts or heapsorts are on the chart, as their results
are significantly worse than the other results, which makes it difficult to
accurately see them.

@@

This section lists the contributions of this report. Each of these is novel  
and was discovered or developed during the course of this research.

The primary contribution of this report is that it repeats the work of LaMarca
on cache-conscious sorting, and validates many of his claims, while questioning
others. The next most important contribution is having performed a branch
prediction analysis of major sorts - which to my knowledge has not been done
before - and developing a framework by which these results can be repeated and
expanded upon.

### 2005-08-30 docs/tech_report/heap.tex
formatting fixes. -- and "\.  "
tidied up

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches will probably be correctly predicted.
An 8-heap has seven comparisons between children, and will probably have about
1.59 mispredictions, as it behaves similarly to selection sort (See \ref{selection
branches harmonics}). It is possible, therefore, that the 8-heap will have slightly
fewer mispredictions due to the combination of the shallower heap and the larger
fan-out.

>>>>

It is expected that the number of comparison branches stays the same with
an increased fanout. When the height of the heap halves, the number of
comparisons in each step doubles. However, the proportion of those which are
predictable changes. The key being sifted down the heap is likely to go the
entire way down the tree. Therefore the comparison between the smallest child
and the parent is predictable. However, the comparisons to decide which child is
smallest are not predictable. In a 2-heap, there is one comparison between
children, meaning that half the branches will probably be correctly predicted.
An 8-heap has seven comparisons between children, and will probably have about
1.59 mispredictions, as it behaves similarly to selection sort (See \ref{selection
branches harmonics}). It is possible, therefore, that the 8-heap will have slightly
fewer mispredictions due to the combination of the shallower heap and the larger
fanout.

### 2005-08-30 docs/tech_report/merge.tex
tidied up

Double aligned multi-mergesort is an even more complicated algorithm. It sorts
the arrays fully in the cache, leaving them bitonically arranged (sorted into
sub-arrays, sorted into alternately increasing and decreasing order). The final
merge is performed separately, and hence no bias for the forward branches, as
there was in tiled mergesort. The final sort is performed with a $k$-way merge,
which is implemented using a heap data structure. This is shown in Figures
\ref{mergesorts-5} and \ref{mergesorts-6}. The `\n{Comparison}' branches are
comparisons from the merge. The labels in these graphs are the same as in Figure
\ref{mergesorts-4} and \ref{base and memory-tuned heapsorts-2}.

@@

Overall, when the $k$-ary heap is used, there are approximately the same total
number of branch mispredictions as in the other mergesort algorithms. Although
the final $k$-way merge executes a lot more comparison branches, these heap
branches are quite predictable, as we saw in Section
\ref{heapsort-branch-prediction}. This means that using the technique to reduce
cache misses does not adversely affect branch prediction rates (it does,
however, still affect the instruction count). The rate of branch misses is much
the same as that in heapsort, and progresses in the same manner (see Figure
\ref{memory-tuned heapsort curve}). Note that `\n{comparison 0}' is included for
illustration purposes only. It is a comparison between two values, one of which
has just been assigned to the other, and which are therefore equal. It is
guaranteed to be taken every time, and would be optimized out by the compiler.

>>>>

Double aligned multi-mergesort is an even more complicated algorithm. It sorts
the arrays fully in the cache, leaving them bitonically arranged (sorted into
sub-arrays, sorted into alternately increasing and decreasing order). The final
merge is performed separately, and hence no bias for the forward branches, as
there was in tiled mergesort. The final sort is performed with a k-way merge,
which is implemented using a heap data structure. This is shown in Figures
\ref{mergesorts-5} and \ref{mergesorts-6}. The `\n{Comparison}' branches are
comparisons from the merge. The labels in these graphs are the same as in Figure
\ref{mergesorts-4} and \ref{base and memory-tuned heapsorts-2}.

@@

Overall, when the $k$-ary heap is used, there are approximately the same total
number of branch mispredictions as in the other mergesort algorithms. Although
the final k-way merge executes a lot more comparison branches, these heap
branches are quite predictable, as we saw in Section
\ref{heapsort-branch-prediction}. This means that using the technique to reduce
cache misses does not adversely affect branch prediction rates (it does,
however, still affect the instruction count). The rate of branch misses is much
the same as that in heapsort, and progresses in the same manner (see Figure
\ref{memory-tuned heapsort curve}). Note that `\n{comparison 0}' is included for
illustration purposes only. It is a comparison between two values, one of which
has just been assigned to the other, and which are therefore equal. It is
guaranteed to be taken every time, and would be optimized out by the compiler.
